---
title: "class12 hw population scale analysis"
author: "amy (pid A16962111)"
format: pdf
editor: visual
---

## population scale analysis

**Q13.** Read this file ([https://bioboot.github.io/bggn213_W19/class](https://bioboot.github.io/bggn213_W19/class-)-material/rs8067378_ENSG00000172057.6.txt) into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.csv")
head(expr)
```

The sample size for each genotype is as follows: A/A has 108 samples, A/G has 233 samples, and G/G has 121 samples.

```{r}
table(expr$geno)
```

The median expression level is 31 for A/A, 25 for A/G, and 20 for G/G.

```{r}
library(dplyr)
expr %>%
  group_by(geno) %>%
  summarize(median(exp))
```

**Q14.** Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

From the boxplot below, we can infer that the relative expression between A/A and G/G is significantly different because there is no overlap between the boxes. Since G is the ancestral allele, we can say that the SNP (allele A) increases expression of ORMDL3.

```{r}
library(ggplot2)

ggplot(expr) +
  aes(geno, exp, fill=geno) +
  geom_boxplot(notch=TRUE)
```
