---
title: "class18: pertussis mini-project"
author: "amy (pid A16962111)"
format: md
editor: visual
---

First we will examine and explire Pertussis case numbers in the US as tracked by the CDC: <https://www.cdc.gov/pertussis/surv-reporting/cases-by-year.html>

We can use the `datapasta()` package to scrape this data from the website into R

```{r}
#install.packages("datapasta")
library(datapasta)
```

```{r}
cdc <- data.frame(data.frame(year = c(1922L,1923L,1924L,1925L,1926L,1927L,1928L,1929L,1930L,
                                      1931L,1932L,1933L,1934L,1935L,1936L,1937L,1938L,1939L,
                                      1940L,1941L,1942L,1943L,1944L,1945L,1946L,1947L,1948L,
                                      1949L,1950L,1951L,1952L,1953L,1954L,1955L,1956L,1957L,
                                      1958L,1959L,1960L,1961L,1962L,1963L,1964L,1965L,1966L,
                                      1967L,1968L,1969L,1970L,1971L,1972L,1973L,1974L,1975L,
                                      1976L,1977L,1978L,1979L,1980L,1981L,1982L,1983L,1984L,
                                      1985L,1986L,1987L,1988L,1989L,1990L,1991L,1992L,1993L,
                                      1994L,1995L,1996L,1997L,1998L,1999L,2000L,2001L,2002L,
                                      2003L,2004L,2005L,2006L,2007L,2008L,2009L,2010L,2011L,
                                      2012L,2013L,2014L,2015L,2016L,2017L,2018L,2019L,2020L,
                                      2021L),
                     cases = c(107473,164191,165418,152003,202210,181411,161799,197371,166914,
                               172559,215343,179135,265269,180518,147237,214652,227319,103188,
                               183866,222202,191383,191890,109873,133792,109860,156517,74715,
                               69479,120718,68687,45030,37129,60886,62786,31732,28295,32148,
                               40005,14809,11468,17749,17135,13005,6799,7717,9718,4810,3285,
                               4249,3036,3287,1759,2402,1738,1010,2177,2063,1623,1730,1248,
                               1895,2463,2276,3589,4195,2823,3450,4157,4570,2719,4083,6586,
                               4617,5137,7796,6564,7405,7298,7867,7580,9771,11647,25827,25616,
                               15632,10454,13278,16858,27550,18719,48277,28639,32971,20762,
                               17972,18975,15609,18617,6124,2116)
                   ))
```

```{r}
head(cdc)
```

I want a plot of cases per year with ggplot

```{r}
library(ggplot2)
ggplot(cdc, aes(year, cases)) +
  geom_line() +
  geom_vline(xintercept = 1947, col="blue") +
  geom_vline(xintercept = 1992, col="red") +
  geom_vline(xintercept = 2020, col="orange")
```

Access data from the CMI-PB project

This database (like many modern projects) uses an API to return JSON format data

We will use the R package `jsonlite()`

```{r}
#install.packages("jsonlite")
library(jsonlite)

subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE) 
head(subject)
```

How many wP (the older whole-cell vaccine) individuals and aP (newer acellular vaccine) individuals are in this data set?

```{r}
table(subject$infancy_vac)
```

What is the number of individuals by biological sex and race?

```{r}
table(subject$race, subject$biological_sex)
```

Use `lubridate()` to calculate the age of everyone in the dataset

```{r}
#install.packages("lubridate")
library(lubridate)

subject$age <- time_length(today() - ymd(subject$year_of_birth), "years")
```

```{r}
ggplot(subject) + 
  aes(age) +
  geom_histogram() +
  facet_wrap("infancy_vac")
```

Get more data from CMI-PB

```{r}
specimen <- read_json("https://www.cmi-pb.org/api/specimen", simplifyVector = TRUE) 
head(specimen)
```

We need to **join** these two tables (subject and specimen) to make a single new "meta" table with all our metadata. We will use the `dplyr()` join functions to do this

```{r}
library(dplyr)
meta <- inner_join(subject, specimen)
head(meta)
```

Now we can read some of the other data from CMI-PB

```{r}
ab_titer <- read_json("https://www.cmi-pb.org/api/v4/plasma_ab_titer", simplifyVector = T)
head(ab_titer)
```

One more `inner_join()` to add all our metadata in `meta` onto our `ab_titer` data table

```{r}
abdata <- inner_join(meta, ab_titer)
head(abdata)
```

```{r}
ggplot(abdata) +
  aes(MFI, antigen) +
  geom_boxplot()
```

Why are certain antigens and not others very variable in their detected levels here?

```{r}
ggplot(abdata) +
  aes(MFI, antigen, col=infancy_vac) +
  geom_boxplot()
```

There are potentially some differences here but in general it is hard to tell with this whole dataset overview

```{r}
table(abdata$dataset)
```

Let's focus in on just the `2021_dataset`, and on PT antigen IgG levels

```{r}
abdata.21 <- filter(abdata, dataset == "2021_dataset")
pt.21 <- filter(abdata.21, isotype == "IgG", antigen == "PT")
```

plot of days relative to boost vs MFI levels

-   **Q17.** Do you see any clear difference in aP vs. wP responses?

    Yes, wP individuals generally have higher antigen levels.

```{r}
plot <- ggplot(pt.21) +
  aes(planned_day_relative_to_boost, MFI_normalised, col=infancy_vac, group=subject_id) +
  geom_point() +
  geom_line()

plot
```
