---
title: "hw class06 r functions"
author: "amy (pid A16962111)"
format: pdf
editor: visual
---

## original code

```{r}
library(bio3d)
s1 <- read.pdb("4AKE") # kinase with drug
s2 <- read.pdb("1AKE") # kinase no drug
s3 <- read.pdb("1E4Y") # kinase with drug
s1.chainA <- trim.pdb(s1, chain="A", elety="CA")
s2.chainA <- trim.pdb(s2, chain="A", elety="CA")
s3.chainA <- trim.pdb(s1, chain="A", elety="CA")
s1.b <- s1.chainA$atom$b
s2.b <- s2.chainA$atom$b
s3.b <- s3.chainA$atom$b
plotb3(s1.b, sse=s1.chainA, typ="l", ylab="Bfactor")
plotb3(s2.b, sse=s2.chainA, typ="l", ylab="Bfactor")
plotb3(s3.b, sse=s3.chainA, typ="l", ylab="Bfactor")
```

## new `protein_drug_interactions_analysis` function

-   Input: any PDB file

-   What it does and how to use it: analyzes protein-drug interactions by creating a plot of Bfactor trends and secondary structure from protein structure data. To use, call the function and enter any PDB code in `""` ; e.g. `protein_drug_interactions_analysis("4AKE")`

-   Output: a lineplot of Bfactor trends with marginal rectangles representing secondary structure

```{r}
protein_drug_interactions_analysis <- function(x) {
  # read protein structure file
  s <- read.pdb(x)
  # select calcium atoms in chain A
  s.chainA <- trim.pdb(s, chain="A", elety="CA")
  # save Bfactor data
  s.b <- s.chainA$atom$b
  # create a lineplot of Bfactor trends with marginal rectangles representing secondary structure
  plotb3(s.b, sse=s.chainA, typ="l", ylab="Bfactor")
}
```

## example output

```{r}
protein_drug_interactions_analysis("4AKE")
```
